---
title: "Exploratory Data Analysis"
author: "Adam Shelton"
date: "5/15/2020"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(skimr)
library(kableExtra)
library(cowpoke)
library(ggcorrplot)
library(beachball)
library(here)

knitr::opts_chunk$set(echo = TRUE, dpi = 200)

print_kable = function(x, caption = "") {
  x %>% kable(caption = caption) %>% kable_styling(c("responsive", "condensed", "striped"), full_width = FALSE)
}
```

## Load Data

```{r}
analysis_data = readRDS(here("cleaned_data.rds")) %>% mutate_if(is.character, factor)
var_cats = read_csv(here("names_to_keep.csv"))
demo_names = filter(var_cats, discrim)$value %>% .[. %in% names(analysis_data)]
outcome_names = filter(var_cats, outcome)$value %>% .[. %in% names(analysis_data)]

no_demo_data = analysis_data %>% select(-all_of(demo_names))
outcome_data = analysis_data %>% select(all_of(outcome_names))
```



## Descriptive Statistics
```{r}
descr_stats = analysis_data %>% skim() %>% partition()

# descr_stats$logical %>% print_kable("Binary Variables")
descr_stats$factor %>% mutate_if(is.numeric, format_number) %>% print_kable("Factor Variables") 
descr_stats$numeric %>% mutate_if(is.numeric, format_number) %>% print_kable("Continuous Variables")
```

## Correlation Matrix
```{r fig.height=20, fig.width=20}
analysis_data %>% mutate_all(as.numeric) %>% cor(use = "pairwise.complete.obs") %>% ggcorrplot(p.mat = cor_pmat(.), hc.order = TRUE)
# p.mat = cor_pmat(.), hc.order = TRUE
```

## Flight Risk Accuracy 

Here we are trying to estimate the "accuracy" of those determining whether a person charged with a crime is a flight risk, and therefore should be detained until their trial. While we obviously cannot know whether someone who was detained until trial would have returned for their trial or not from this data, we can attempt to measure this from those who were released after paying their bail. Someone who was released on bail would likely have been considered a significantly greater flight risk than a person released with no bail. We can attempt to measure the amount of false positives by looking at people who were deemed a high enough flight risk for bail, but still returned. Those who were released on bail and did not return are considered to be "correctly" classified here. If someone was released and returned then they were correctly classified, while those who were released without bail and did not return would be considered a false negative. This is not a perfect measurement of classification accuracy, since bail is a variable amount, and is different than being released even if a high flight risk with no collateral, but should be close enough to determine how accuracte and equitable humans are in this process.

```{r}
tibble(" " = c("Paid Bail", "No Bail"), "Appeared at Court" = c("False Positive", "Correct"), "Did Not Appear" = c("Correct", "False Negative")) %>% print_kable()

accuracy_data = analysis_data %>% 
  filter(RELDET1 == "Released") %>% 
  filter(!is.na(FTA1)) %>%
  mutate(no_bail = (FINREL == "Not made bail"), 
         fta = FTA1 == "Yes, FTA", 
         pred_acc = ifelse(fta, 
                           ifelse(no_bail, "False Negative", "Correct"), 
                           ifelse(no_bail, "Correct", "False Positive")),
         race = ifelse(WHITE == "Yes", "White", ifelse(BLACK == "Yes", "Black", ifelse(HISP == "Yes", "Latinx", "Other"))))

# accuracy_data %>% select(FINREL, no_bail, fta, FTA1, pred_acc) %>% distinct() %>% View()

accuracy_data %>% count(pred_acc) %>% na.omit() %>% mutate(percent = scales::percent(n / sum(n), accuracy = 0.1)) %>% select(-n) %>% rename("Prediction Accuracy" = pred_acc) %>% print_kable("Overall Accuracy")

accuracy_data %>% count(GENDER, pred_acc) %>% na.omit() %>% crosstab_percent(formatted = TRUE) %>% select(-n) %>% pivot_wider(names_from = pred_acc, values_from = percent) %>% rename("Gender" = GENDER) %>% print_kable("Accuracy by Gender")

accuracy_data %>% count(AGED, pred_acc) %>% na.omit() %>% crosstab_percent(formatted = TRUE) %>% select(-n) %>% pivot_wider(names_from = pred_acc, values_from = percent) %>% rename("Age Category" = AGED) %>% print_kable("Accuracy by Age Category")

accuracy_data %>% count(race, pred_acc) %>% na.omit() %>% crosstab_percent(formatted = TRUE) %>% select(-n) %>% pivot_wider(names_from = pred_acc, values_from = percent) %>% rename("Race" = race) %>% print_kable("Accuracy by Race")
```

## Equity Tables
```{r}
equity_table = function(data_set, demo_var, recid_var, demo_label = str_to_sentence(demo_var), recid_label = "Rate", format_perc = TRUE) {
  data_set %>% 
    count(!!sym(demo_var), !!sym(recid_var)) %>% 
    na.omit() %>% 
    crosstab_percent(vars = demo_var, formatted = format_perc) %>% 
    select(-n) %>% 
    pivot_wider(names_from = recid_var, values_from = "percent") %>% 
    select(-`FALSE`) %>% (function(x) {names(x) = c(demo_label, recid_label); x})
}

equity_data = analysis_data %>% 
  mutate(no_bail = (FINREL == "Not made bail"), 
         fta = FTA1 == "Yes, FTA", 
         race = ifelse(WHITE == "Yes", "White", ifelse(BLACK == "Yes", "Black", ifelse(HISP == "Yes", "Latinx", "Other"))))
```

### Release Rates
```{r}
equity_data %>% 
  mutate(RELDET1 = RELDET1 == "Released") %>% 
  count(RELDET1) %>% 
  na.omit() %>% 
  mutate(Percent = scales::percent(n / sum(n), accuracy = 0.1), RELDET1 = ifelse(RELDET1, "Yes", "No")) %>%
  rename("Released" = RELDET1) %>% 
  select(-n) %>% 
  print_kable("Overall Release Rate")

equity_data %>% 
  mutate(RELDET1 = RELDET1 == "Released") %>% 
  equity_table("GENDER", recid_var = "RELDET1") %>% 
  print_kable("Release Rate by Gender")

equity_data %>% 
  mutate(RELDET1 = RELDET1 == "Released") %>%  
  equity_table("AGED", recid_var = "RELDET1", demo_label = "Age Group") %>% 
  print_kable("Release Rate by Age Group")

equity_data %>% 
  mutate(RELDET1 = RELDET1 == "Released") %>% 
  equity_table("race", recid_var = "RELDET1") %>% 
 print_kable("Release Rate by Race")
```
### Bail Rates
```{r}
bail_data = equity_data %>% filter(RELDET2 == "Financial release" | RELDET2 == "Held on bail" | RELDET2 == "Denied bail") %>% mutate(offered_bail = RELDET2 != "Denied bail")
```

On average, `r bail_data$offered_bail %>% mean() %>% scales::percent(accuracy = 0.1)` of people charged with a crime and were not released, were offered to be released on bail.

```{r}
bail_data %>% filter(!is.na(GENDER)) %>% group_by(GENDER) %>% summarise("Rate" = mean(offered_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Gender" = GENDER) %>% print_kable("Bail Rate by Gender")

bail_data %>% filter(!is.na(AGED)) %>% group_by(AGED) %>% summarise("Rate" = mean(offered_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Age Category" = AGED) %>% print_kable("Bail Rate by Age Category")

bail_data %>% filter(!is.na(race)) %>% group_by(race) %>% summarise("Rate" = mean(offered_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Race" = race) %>% print_kable("Bail Rate by Race")
```


### Bail Amount
```{r}
bail_amt_data = equity_data %>% filter(!(is.na(BAILAMT)))
```

The median bail amount overall is \$`r bail_amt_data$BAILAMT %>% median() %>% format_number()`.
```{r}
bail_amt_data %>% select(GENDER, BAILAMT) %>% na.omit() %>% group_by(GENDER) %>% summarise(BAILAMT = median(BAILAMT)) %>% rename("Gender" = GENDER, "Median Bail" = BAILAMT) %>% mutate_if(is.numeric, format_number) %>% print_kable("Median Bail Amount by Gender")

bail_amt_data %>% select(AGED, BAILAMT) %>% na.omit() %>% group_by(AGED) %>% summarise(BAILAMT = median(BAILAMT)) %>% rename("Age Category" = AGED, "Median Bail" = BAILAMT) %>% mutate_if(is.numeric, format_number) %>% print_kable("Median Bail Amount by Age Category")

bail_amt_data %>% select(race, BAILAMT) %>% na.omit() %>% group_by(race) %>% summarise(BAILAMT = median(BAILAMT)) %>% rename("Race" = race, "Median Bail" = BAILAMT) %>% mutate_if(is.numeric, format_number) %>% print_kable("Median Bail Amount by Race")
```

### Paid Bail Rates

```{r}
paid_bail_data = bail_amt_data %>% mutate(paid_bail = FINREL == "Made bail") %>% filter(!is.na(paid_bail))
```

On average, `r paid_bail_data$paid_bail %>% mean() %>% scales::percent(accuracy = 0.1)` of people charged with a crime were able to pay their bail.

```{r}
paid_bail_data %>% filter(!is.na(GENDER)) %>% group_by(GENDER) %>% summarise("Rate" = mean(paid_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Gender" = GENDER) %>% print_kable("Paid Bail Rate by Gender")

paid_bail_data %>% filter(!is.na(AGED)) %>% group_by(AGED) %>% summarise("Rate" = mean(paid_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Age Category" = AGED) %>% print_kable("Paid Bail Rate by Age Category")

paid_bail_data %>% filter(!is.na(race)) %>% group_by(race) %>% summarise("Rate" = mean(paid_bail) %>% scales::percent(accuracy = 0.1)) %>% rename("Race" = race) %>% print_kable("Paid Bail Rate by Race")
```


